### 2019-11-20 rebuttal/rebuttal.tex

% link packages
\usepackage{url}
\usepackage{doi}

%%%%%%%%%%%%%%%%%%%%%
%%% DEFINE STYLES %%%
%%%%%%%%%%%%%%%%%%%%%

% set font after chapter sonny style
\usepackage{fontspec}
\setmainfont [Path = ../fonts/,
UprightFont = *-300,
ItalicFont = *-300-Italic,
BoldFont = *-700,
BoldItalicFont = *-700-Italic
]{MuseoSans}

\hypersetup{
	colorlinks = true,
%	linkcolor = {blue},
  urlcolor = {blue},
	citecolor = {med-gray}
}

\newcommand{\exam}[2][\  ]{\hspace{0pt}\marginpar{\color{red}#1}$\bullet$ \textit{#2}}
\newcommand{\imp}[1]{{\color{red} #1}}
\newcommand{\nimp}[1]{{\color{gray} #1}}
\newcommand{\answ}[1]{{\color{blue} $\triangleright$ #1}}

\usepackage{tikz}
\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
		\node[shape=circle,draw,inner sep=2pt] (char) {#1};}}

@@

\newcommand{\bigexclaim}{\raisebox{-0.1em}{\BigTriangleUp}\hspace{-0.32em}\llap{\small\textbf{!}}\hspace{0.32em}}
\newcommand{\bigquest}{\circled{?}}

\usepackage{fdsymbol}
\newcommand{\tagimp}{\bigexclaim}
\newcommand{\tagtime}{{\Large $\hourglass$}}

\begin{document}

\section{Summary of most important comments, per chapter}

\textbf{Overall:} The quality of writing, layout and graphical presentation is of a lower quality in the non-published parts of the thesis. Certain unpublished chapters feel rushed and unfinished. Furthermore, for some chapters it feels that they could be merged into a broader discussion chapter.

\textbf{Overall:} Merge chapters 8-9-10 into one discussion.

\textbf{Overall:} Merge chapter 8 into chapter 2.

\textbf{Overall:} Random Forest variable importances are often used to detect differentially expressed genes. Why RF and not more standard statistical tests?

\textbf{Introduction:} {Describe the practical implications of the
		technological advances (e.g. precision medicine, new treatments) and
		motivate the research from a non-technical view point (How is/can this PhD
		change the world?)

\textbf{dyngen:} {Table 2.2: explain default values and formulae}

\textbf{dyngen:} {How crucial are certain parts of the GRN model, e.g. mRNA production and degradation?}

\textbf{dyngen:} {Compare against other scRNA-seq generators}

\textbf{dyngen:} {It is claimed dyngen is a multi-modality simulator, yet it produces only transcriptomics data}

\textbf{dynbenchmark:} {Comparisons focus mainly on relative differences between methods. It would have been interesting to also assess the absolute performance of each method.}

\textbf{dynbenchmark:} {No statistical tests were performed to assess whether methods are significantly different from one another}

\textbf{dynbenchmark:} Turn into UK English to be consistent with the rest of the dissertation.

\textbf{dyno:} {Does not really read as a scientific paper (yet).}

\textbf{SCORPIUS:} {The methodology has changed between the bioRxiv preprint and this chapter, yet Figure 5.4 is exactly the same. Did the changes have no (measurable) effect on this analysis?}

\textbf{SCORPIUS:} {Discuss the computational complexity of SCORPIUS.}

\textbf{bred:} {Better explain how bred is different from GENIE3.}

\textbf{bred:} {What makes bred a true case-wise NI method? Provide a more formal definition of what a case-wise regulatory network is.}

\textbf{bred:} {Compare the bred clustering against regular clustering obtained from expression profiles}

\textbf{bred:} {Compare against other methods, e.g. GENIE3 + regulon specificity score}

\textbf{bred:} {Discussion is very short, please elaborate}

\textbf{IncGraph:} {On optimising GRNs, compare a simple MST with IncGraph}

\textbf{Benchmarking guidelines:} {As the candidate is not first author on this paper, I’m not sure if this paper
	should be included in the PhD book verbatim.}

\textbf{Discussion:} {The Discussion chapter is missing a more broad context
			description. How are other researchers using these tools to gain new biological insights? Where is the field of modelling single-cell dynamics going?}

\textbf{Discussion:} {Discuss the links between different chapters}

\newpage
\section{Legend}
\exam{Question or comment by jury member. \nimp{Description of dissertation.}}

\exam[\tagimp]{\imp{Important matter} $\rightarrow$}

\exam[\tagtime]{Matter which will take some time to reply to / fix $\rightarrow$}

\answ{Response or statement by candidate.}

@@

\section{Joeri Ruyssinck}

\answ{Small changes and typos have been fixed, unless noted otherwise.}

@@

\subsection{Strengths of the thesis}

\exam{\nimp{The research domain of this thesis is young and technical evolutions in the
	encompassing field move fast. This makes it especially challenging to do
	research in the field, as often the original PhD thesis subject is obsolete at the
	end. The candidate has done exceptionally well to not only suggest and create
	novel algorithms which are at the state-of-the-art but also to grasp and tackle
	the greater high level challenges in the field, resulting in a potential seminal
	benchmarking paper.}}

\exam{\nimp{The thesis as a whole contains many relevant contributions, of which several
	have been published in high quality journals.}}

\exam{\nimp{The figures in the thesis are of exceptional quality. They are creative and able
	to convey large volumes of information to the reader in a visually pleasing
	format.}}

\exam{\nimp{The introduction chapter of the thesis demonstrates that the candidate
	understands and is able to convey the background of both (computational)
	biology and computer science.}}

\exam{\nimp{The passion of the candidate to bring modern good software development
	practices to the field is present throughout the work and might contribute to a
	more mature algorithm development in computational biology.}}

@@

\subsection{Weaknesses of the thesis}
\exam[\tagimp \tagtime]{\imp{The quality of writing, layout and graphical presentation is of a lower quality in the non-published parts of the thesis. Certain unpublished chapters feel rushed and unfinished. Furthermore, for some chapters it feels that they could be merged into a broader discussion chapter.}}

\exam[\tagimp \tagtime]{\imp{The Introduction and Discussion chapter is missing a more broad context
	description. After reading the work, I gained very little insight how other
	researchers are using these tools to advance science.}}

\exam{\nimp{The main contribution of this work is the excellent benchmarking study
	published in Nature Biotechnology.} However, this also effects the overall
	cohesiveness of the work, as information is repeated throughout the book at
	several levels of detail. \nimp{This however a common side-effect of grouping
	published manuscripts into a thesis and I’m aware there is no (quick) solution
	to this.}}

\exam[\tagimp]{\nimp{It is very rare to see that research groups (or individual researchers) invest this
	much effort and time in creating qualitative software. I agree with the ideas of
	the candidates that these practices to adopt software to accompany research
	ideas should be encouraged.} Yet I wonder how the current academic system
	can support such concepts. Academics often hold different positions in
	different research groups around the world, who should be the maintainer in
	these cases? The original PhD students that developed the work? The original
	research group? Will such higher standards not lead to barriers which might
	be harder to overcome for PhD students which might be constrained by limited
	support or funding?}

@@

\subsection{Summary and chapter 1}

\exam[\tagimp \tagtime]{\nimp{As an interdisciplinary thesis, it is very much enjoyable that the introduction
	chapter contains a concise, yet precise overview of the concepts which are
	needed to understand the work of the candidate. I believe the chapter is nicely
	written and especially the flow of information is well chosen. However, the
	candidate only briefly touches upon the societal value of the technological
	innovations he describes. For a PhD thesis, } \imp{I think it would be an added value
	if the candidate could briefly describe the practical implications of the
	technological advances (e.g. precision medicine, new treatments) and
	motivate the research from a non-technical view point (How is/can this PhD
	change the world?)}}

\exam{Figure 1.9 does not contain SCORPIUS (probably published before,
	but still a bit weird for a PhD containing the manuscript)} \\
\answ{Indeed, this figure was adapted from Cannoodt et al. 2016 and serves as an illustration that
TI methods often have common building blocks. Extending this figure to >75 TI methods would decrease the interpretability of the figure.}

\exam{Bold is only used in 1.1.2 and research challenges} \\
\answ{Usage of the bold font style in 1.1.2 and 1.3 serve clear, but different, purposes. In 1.1.2, three words are highlighted to establish a link between each term's introduction in the first paragraph, and their explanations in the second paragraph. In 1.3, important sections of each research objective are highlighted in order to highlight the structure of this section.}

\exam{Use of we/I in summary} \\
\answ{Usage of first person pronouns in this dissertation is very limited. In the summary of my contributions to the field, I believe usage of the first person improves legibility in comparison to when the phrases in question are converted to a passive form.}

\subsection{Chapter 2}

\exam{It is not clear to me what is meant with ‘such as sampling a cell at a certain
	time point and once more at a later point’ and why this is currently not
	possible?}

\exam{Figure 2.1.A. What does the figure in ‘Combine simulations’ represent? Are
		these single simulations which are mapped continuously to the backbone?}

\exam{Figure 2.5. It is not explained what Subfigure F represents. Only after the
	reader has completed the chapter, it becomes clear that the red lines probably
	represent snapshots and the lines, curves of gene expression.}

\exam{Figure 2.6: Consider changing the dashed lines, as they are not (clearly)
		visible printed}

\exam{Figure 2.6 and the example text for the cyclical example is unclear.}

\exam{While I
		understand that the instructions result in the desired state change, I also see
		many other configurations which would result in a cyclical example.}

\exam{Why is it currently needed to explicitly keep certain module expressions constant to
		determine the backbone? Would the same result not be achieved by simply
		simulating X times and averaging expression levels?}

\exam{Figure 2.7. The figure legend could be extended}

\exam{The introduction of FANTOM5 is abrupt and it would be better if the context
		would be briefly explained to the reader.}

\exam{The paragraph describing Step 3 is unclear: -> Target genes ARE? Regulated
		... but is-> are ?}

\exam{Figure 2.8 can incorrectly suggest that target genes can only have a single
		regulator assigned}

\exam[\tagimp \tagtime]{\imp{Table 2.2. contains many default ‘magic’ numbers?} It would be nice to know
		how in practice you ended up selecting them and which problems you
		encountered during design.}

\exam{Since these are default values, it would suggest
		that the user can modify them? Why would a user do that and what would be
		the impact?}

\exam{Am I correct that currently splicing offers little extra value as the formula’s
		would imply that alternative splicing is not supported?}

@@

\exam[\tagimp \tagtime]{\imp{During a first read, it was confusing at times to read specifically about the
	strategy to determine the backbone while it had not been described how the
	simulation was performed.} At a certain time I was under the impression that
	dyngen did not simulate cell evolution end-to-end but only between states.
	Would the readability be improved if the order was changed in which the
	concepts are explained? The backbone is only needed for the visualisation.}

\exam{2.4.6. is a bit worrying. Could I reproduce dyngen without knowing these
		details? Does the sampling process not deserve more explanation?}

@@

\subsection{Chapter 3}

\exam{There are some artefacts of converting the paper into a PhD chapter: e.g.
	references to Supplementary tables/figures should be changed to the actual
	table number.} \\
\answ{Certain supplementary figures and files had not been included in the dissertation, but instead the reference contained a clickable URL which directs the reader to a downloadable file. Since this information is lost to readers of the printed dissertation, explicit captions have been added in a section for supplementary figures and tables, and contain a printable URL which direct the reader towards the file.}

\exam{Some words are marked with an asterisk but the footnote? is
	missing.} \\
\answ{Section 3.4.2 uses an asterisk to denote special cases and does not refer to a missing footnote. This notation is introduced in the paragraph leading up to the usage thereof. An actual asterisk has been added to the introduction of the notation to improve readability of this section.}

\exam{You mention consensus predictions, is this something you investigated more
		in depth? Would ‘ensemble’ predictions result in better performance? Is there
		a good way to aggregate the predictions?}

\exam{Monocle DDRTree seems to overestimate the topology based on Figure 3.5, is
	this somehow related to the way you have to force the algorithm into the
	common format?}

\exam{The dyngen described here seems to be less advanced than the one from
		Chapter 2?}

\exam[\tagtime]{Consistency: this chapter is US English}

@@

\subsection{Chapter 4}

\exam{Figure 4.7: The figure is lacking what the colors represent.}

\exam{Figure 4.7: I see that even if you use the same dimensionality
	reduction; there is still a lot of variation between the methods on a very simple
	example. Could you clarify
	how often these methods disagree on trivial examples?}

\exam{Dyno offers a lot of
	guidance with respect to how to choose your methods, run them, etc. Did you
	consider support for features such as interpretation or automated error
	detection after prediction?}

@@

\exam[\tagimp \tagtime]{\imp{This chapter does not really read as a scientific paper (yet).} There are few
	comments I have since this is mainly an instruction manual + feature overview.}

\exam{Figure 4.7 is missing slingshot and paga label in part B} \\
\answ{Figure 4.7 does contain labels for Slingshot and PAGA. A small gap between part A and part B has been added, to improve interpretability.}

@@

\subsection{Chapter 5}

\exam[\tagimp]{Comparing the preprint with the chapter in the thesis. Some minor changes
	seem to have been made to SCORPIUS: e.g. I believe in the pre-print
	classical Torgerson MDS is mentioned, while the more optimized Landmark
	MDS is used in the chapter and final version. There is also no mention of
	outlier removal. The main result Figure 5.4 is however identical. \imp{Did this
	change or other potential changes have no (measurable) effect on the initial
	analysis? I also noticed that the biological conclusions or hypothesis have
	been reduced in claim strength in the paper or have been removed.} It would
	be interesting to know if this is due to new insights that have been found in the
	period 2016-> 2019 or mainly because the paper’s target audience has
	shifted.} \\

\answ{As mentioned, the preprint was originally published in 2016. In order to submit this work to a peer-reviewed journal in 2019, the manuscript and the software was rewritten to reflect developments within the field.

@@

Regarding the methodology. Classical MDS has indeed been replaced with Landmark MDS, since this allows scaling beyond 10'000 cells. Principal curves has been modified to also allow scaling beyond 10'000 cells. These changes have little to no effect on any results. Outlier removal was removed because this step is now part of mainstream preprocessing pipelines such as Seurat. Initialisation of the curve (k-means clustering + shortest path) has been removed -- while this part of the algorithm showed improved performance when applied to the datasets included in the bioRxiv benchmark, this was not the case when applied to a larger set of datasets.

Regarding reducing the biological conclusions. Section 5.2.2 regarding the functional modules found by SCORPIUS was reduced in character count, yet it makes the same claims. The main differences are that gene names and explanations of the biological functions are not discussed in detail, but are instead summarised in figure 5.4. The reasoning for these changes are indeed to adapt the text to the target audience.
For example, detailed explanation of what actin polymerisation is, was removed; because immunologist readers already know this information and data scientist readers are presented with sufficient information when told that particular genes recovered by SCORPIUS are related to a biological process that is highly relevant to dendritic cell development, that is, actin polymerisation.}

\exam{This chapter would offer information on how the developed algorithms
	are being used to discover new insights, so I’m unsure if it is beneficial if this
	chapter evolves more into a technical description of Scorpius as the former is
	missing from the thesis.}
\answ{???}

\exam{It would help to list the title of the manuscript on the pre-print server} \\
\answ{I fixed the DOI. This should be sufficient to find back the manuscript on the pre-print server.}

@@

\subsection{Chapter 6}

\exam{The abstract seems to be missing a sentence.}

\exam{\imp{A main challenge in network inference has always been choosing a suitable
		cut-off to determine if something is interacting or not.} There are examples of
		regions in the network which are accurate but come at a much later stage in
		the ranking, at which time many spurious edges have been added to other
		parts of the network. I can only assume that this problem becomes even more
		challenging if one wishes to create case-wise GRN.}

\exam{The added value of the last paragraph in 6.3. seems small}

\exam{Why is Scorpius used to generate Figure 6.3? Is this not confusing, since I
		assume we are not working with single cell data?}

\exam{I’m afraid to ask, but I don’t get the reference to the name bred.}

@@

\exam{\imp{I don’t understand how the approach in figure 6.5 is different from GENIE3 and
		how it results in a tensor instead of matrix/ranking of ‘shuffle’ variable
		importance scores.} I believe I’m missing important information on how
		SCENIC works to be able to understand the modifications. I really struggle
		with this chapter: is the method executed for each profile or for all profiles
		together? Do you not end up with a single GRN for each of the profiles and as
		such why would bred be a true case-wise NI method and the Scenic approach
		would not? It would be nice to discuss this more during the defense.}

\exam{The readability of Figure 6.1 could be improved by using a black font}

\subsection{Chapter 8}
\exam{Does this contribution not better fit in an overall discussion of the thesis?}

\subsection{Chapter 9}
\exam[\tagimp]{\imp{As the candidate is not first author on this paper, I’m not sure if this paper
		should be included in the PhD book verbatim.} Perhaps it should not be
		included or it could be merged with chapters 8-9-10 into a single discussion
		chapter. Maybe at the very least, the candidate should state a bit more
		explicitly why the content was included in the PhD book and what the personal
		contribution was.}

\exam{Author contributions is not correct (CS)} \\
\answ{???}

\subsection{Chapter 10}

\exam{\nimp{The chapter contains a concise recap and critical view on the contributions in
		literature.} It however does not discuss Chapter 9 (see comment above).}

\exam{\imp{The discussion could be improved by stating how the tools are allowing new
		biological insights.} Only a small hint is provided in the last sentence of 10.4.}

\section{Pierre Geurts}

\subsection{Remarks alongside summaries}
\exam{Chapter 1 does not go into the technical details of the computational tools, but it gives a gentle and nicely written introduction to the field.}

\exam{The results
	section of Chapter 2 is however limited. I would have appreciated to see more experiments to illustrate the
	quality of the generated data and also to have a comparison with other simulators from the
	literature. But to be fair, it is not clear how data simulators could be compared to each other or
	even how the quality of a given simulator could be quantitatively evaluated.}

\exam{Chapter 4 is a description of the features offered by the toolkit, more than a technical presentation
	or an evaluation of the tools, but more details about every step can be found in the other papers.}

\exam{Regarding Chapter 6: Further experiments
	would be needed to really evaluate the performance of the method, notably on artificial datasets
	where the true case-wise networks would be known, but also on real single-cell omics dataset,
	which seemed to be the original objective of the authors.}

\subsection{Questions / remarks}

### 2019-11-20 rebuttal/rebuttal.tex

\exam[\tagimp]{The focus of the comparison in Paper 2 is on ranking the different methods. This is extremely
	useful obviously. \imp{I think it would have been interesting to take this opportunity to also assess
	the methods on a more absolute scale.} I don’t think the paper provides answer to such
	questions for example: Is a given method good enough, in general or for a given type of
	trajectory? Is the problem of inferring a given type of trajectory solved in a satisfactory way?
	On which kind of trajectories should the research community focus? What are the cases that
	no method can tackle well?}

\exam{Similarly, in Paper 4, the SCORPIUS method is only assessed relatively to other methods.
	There is no doubt that SCORPIUS is a very good method but I’m missing an absolute
	assessment on how well it performs. It would have been interesting to report the correlation
	between the predicted and true pseudo-time values, in addition to the more general score of
	Paper 2 and to show datasets where it performs poorly, if any.}

\exam[\tagimp]{\imp{In Paper 2, I was surprised that no statistical test was performed to assess whether methods
	are statistically different from one another} (or better than random), as this is commonly done
	in machine learning for example. In addition to method ranking, it would have been
	interesting to be able to highlight groups of methods that are not statistically different from
	each other in terms of accuracy.}

\exam[\tagimp]{In several papers, the authors use Random Forests variable importances to find differentially
	expressed genes (along a trajectory, between branches, at some branching point, etc). \imp{I’m
	wondering why they use RF instead of more standard statistical tests for finding these genes.}
	RF are useful when some multivariate effects need to be found but I’m not sure it is the case
	in all situations where RF is used in the thesis. In Paper 2 for example, it is said that mtry is
	set to 1\% to make sure that highly correlated features are not suppressed in the ranking. In
	Paper 4, Figure 5.5, there is a mention that “shallow” random forests are used, which suggest
	that they don’t want to find complex interaction effects. What is the expected benefit of
	using RF instead of statistical tests?}

\exam[\tagimp]{In Paper 4, \imp{I’m missing a discussion of the computational complexity of the SCORPIUS
	method.} In Section 5.2, the authors say that MDS and principal curves have been adapted so
	that they scale linearly but given that the approximated principal curves algorithm is
	initialized by k-means, isn’t the algorithm quadratic? This is confirmed by Figure 3.7 in
	Paper 2 that shows that SCORPIUS scales quadratically with the number of cells.}

\exam[\tagimp]{In Paper 5, it is argued that most case-wise network inference methods consist in post-
	processing a static GRN, unlike the bred method which is a true case-wise NI method. My
	feeling is that bred is not that different in spirit from existing post-processing methods, as it
	starts from a model trained on the full dataset of samples, which is trying to find regulators
	that are relevant for most of the samples. \imp{What makes bred a true case-wise NI method?}
	More fundamentally, \imp{I’m missing a more formal definition of what a case-wise regulatory
	network is.} A network can obviously not be derived from a single sample (as interactions
	between genes cannot be retrieved from a single static expression vector).}

\exam[\tagimp \tagtime]{As mentioned in the conclusion of Paper 5, the bred method needs to be assessed on more
	problems and against ground truth networks. One baseline experiment that I would have
	been interest in at this stage is \imp{a comparison of the clustering in Figure 6.1 obtained from
	the regulomes with a clustering obtained from the original expression profiles.} Does using
	the regulomes really make a difference in terms of the final clustering?}

\exam[\tagtime]{In Paper 7, the improvement obtained with IncGraph on the two gene networks is interesting.
	However, \imp{I would have liked to see a comparison with a very simple heuristic that goes
	through the ranked list of edges and only adds a new edge if it is not redundant with a
	previously added edge} (as in minimum spanning tree construction algorithms).}

\subsection{Suggested modifications}
\answ{Small suggestions have been accepted, unless noted otherwise.}

\exam{During my first read of the thesis, I had difficulty understanding what the author meant by
	“milestone” and by “region of delayed commitment”. Maybe introduce a definition of these
	notions somewhere in Chapter 1, if it fits.}

\exam{I don’t understand footnote 2 on page 108 explaining why the method in Chapter 4 is called
	‘bred’. Maybe clarify this (sorry if I miss something obvious).}

\exam{The title of Chapter 6 is misleading as the method is actually not applied to single-cell data.
	I would suggest to maybe change it into “inferring case-wise regulatory networks”.}

\exam{In Paper 5, Section 6.4.1: how is the minimum threshold on average importance set? This is
	important as this threshold conditions the number of interactions found.}

\exam{Page 5, the sentence “IHC realizes the visualization...to specific antigens” seems to be
	redundant with the next one.}

\section{Stein Aerts}

\exam{The introduction is relatively brief, but well
written.}

\exam{The methods of Chapter 2 are extensive, but the results are very brief. I’m wondering whether
some parts that are now in the Methods section would not fit better in the Results section. Testing how
representative the generated data are is of course a huge challenge, this could have been addressed
perhaps in the Discussion.}

\exam[\tagimp \tagtime]{\imp{It would also be interesting te evaluate the impact of the gene regulatory
models in the simulation. How crucial is it to use such an extensive model of mRNA production and
degradation, and how do these parameters affect the simulated single-cell data?}}

\exam{Chapter 4 is a bit opposite
of chapter 2 - here the results are described, but without methods. For example, how are features
(genes) selected on the branch points, in a branch, etc (which methods are used for this).}

\exam{The marker gene idenfitication is very interesting, but is not described in detail in the Methods, how
is this RF built, and trained? Is it RF regression or classification? What is the topology, why use a RF,
has it been compared to other techniques?}

\exam{In the benchmark of SCORPIUS compared to other
methods, it is mentioned that many data sets are used, but there are no details on which data sets.}

\exam[\tagimp \tagtime]{\imp{In Chapter 6, it would have been interesting
to compare the results of bred with such methods (for example GENIE3 + regulon specificity score).}}

\exam[\tagimp \tagtime]{\imp{The methods could be a little bit more clear on the specific differences between bred and GENIE3.}}

\exam{In Chapter 7, it may have
been interesting to discuss the relationship with the classical network motifs from Uri Alon.}

\exam{Chapter 9 could have
contained a bit more in-depth discussion about trajectory inference in general, biology, and discussion
of the results in the context of other work.}

\section{Vanessa Vermeirssen}

\subsection{Chapter 1}

\exam{\nimp{The introduction is concise and to the point. This is also the case for the introduction sections of the
		different paper chapters. Therefore,} some sections could have been more elaborated on e.g. briefly
		describe and include the differences between the different dimensionality reduction methods and
		define different network inference approaches. \nimp{At the end of the introduction, the objectives are
		clearly put into the research context.}}

\subsection{Chapter 2}

\exam[\tagimp \tagtime]{In the introduction the PhD candidate mentions \imp{other scRNA-seq profile generators i.e. the
state of the art, but in this chapter no comparison is made with those in order to analyze how
well dyngen performs on the generation of scRNA-seq profiles.} In the introduction I also lack
a short description of the workflows/underlying algorithms of those scRNA-seq profile
generators as opposed to that of dyngen. Can you also comment on how dyngen differs from
simulators of gene regulatory networks such as GeneNetWeaver and Neural Gene Network
Constructor?}

\exam[\tagimp]{Dyngen is put forward as a multi-modality simulator as opposed to other scRNA-seq profile
generators. \imp{This multi-modality is mostly reflected in the results as different simulations of
experiments and extracted data, all based on transcriptomics, whereas in the introduction
this multi-modality is defined as the profiling of another data type than scRNA-seq.} Please
clarify this in this chapter.}

\exam{p. 26 The backbone is defined as a network of cellular states. The connection between gene
modules and those cellular states could be made more explicitly clear. Do I understand
correctly that in a specific cellular state specific gene modules of the module network
become/are active?}

\exam{A legend is missing for Fig. 2.7}

\exam{p. 27 Step 1: “from the module IT belongs to”.} \\
\answ{This has been fixed.}

\exam{Write more clearly in step 2: “These
interactions...” Specify.}

\exam{And step 3:”Target genes...” Verb missing in sentence.} \\
\answ{???}

\exam{p.37 Last part missing in sentence? “The overall score between the different dataset sources
	was moderately to highly correlated ...”} \\
\answ{???}

\exam{dyngen is somewhat differently explained here in the Methods section. Please integrate with
	paper 1 chapter and remove redundancy.}

\exam{It is recommended to clarify and extend the discussion on future improvements that could be
	added to the tool, e.g. alignment and differential expression. What is the biological
	significance? What is meant by feature importance tools/score?}

\exam{Is this really the first algorithm for inferring case-wise regulatory networks?}

\exam[\tagimp]{p. 107 Delete in Results: We generate.}

\exam{\nimp{p. 108 “While accuracy of the network of predicted regulatory interactions is lower in
	comparison to experimental validation techniques, network inference methods offer an
	unbiased and high-throughput insight into the regulatory dynamics of a biological system.”} I
	do not agree. Network inference methods do have a bias, depending on the underlying
	algorithm (cfr Marbach et al. 2012, ...), but so do specific experimental techniques that
	measure only an aspect of gene regulation. It is true that precision-recall for network
	inference predictions is not high, so they can only be viewed as hypothesis generators, but
	are nevertheless valid since they offer, relatively easy, a genome-wide view on gene
	regulation.}

\exam{There are significant differences between the methods of Kuijer and Liu, please describe
	these methods in more detail in the introduction.}

\exam{What about case-specific random forests like in the R package ranger, csfr() ?}

\exam[\tagimp]{The sentence “As such, these methods will most likely recover the interactions that are
	prevalent in the whole population, and will miss interactions that are specific to only a sub-
	population” is incorrect. \imp{I would suggest you read the relevant method papers again, see
	above.}}

\exam{p. 114, 6.4.3. Can you include a motivation for the specific clustering algorithms used here?}

\exam[\tagimp \tagtime]{\imp{The discussion is very short, please elaborate on}: possible applications in the field of
	personalized precision medicine and comparison to state of the art.}

\subsection{Chapter 7}

\exam{Please define graphlet and orbit more precisely in the introduction and differentiate between
	the concepts network motif, subgraph and graphlet.}

\exam{Can you also brieflly introduce the non-incremental approach Orca?}

\subsection{Chapter 8}

\exam{A large part of this chapter is already incorporated in previous chapters, especially in paper2.
	I would suggest to integrate both and avoid redundancy. A large part of this chapter would fit
	in the discussion of paper2.}

\subsection{Chapter 10}
\exam[\tagimp \tagtime]{\imp{The general discussion on the thesis is only 3 pages and consists of small, but critical, discussion
		paragraphs on each of the different papers. The general discussion is too brief and lacks future perspectives on the field. It provides some links between the different chapters,
		but this bridging could be increased further.
		I miss an overarching discussion on where the field of modelling single cell dynamics is going.}}

### 2019-11-20 ch10_discussion/content.tex
update

We developed \texttt{dyngen}, a simulator of single cells to benchmark and stress-test computational tools for single-cell omics (Chapter~\ref{chap:dyngen}).
The \texttt{dyngen} software supports a selection of existing experimental protocols, and can also simulate experiments that are possible with the current technologies, such as generating omics profiles from the same cell at multiple timepoints.
The generated data can help kick-start emerging domains with low data availability more safely by allowing software developers to test their method before numerous datasets become publicly available.

@@

The topological 'profile' of a network was defined as the frequencies of particular topological patterns called 'graphlets'. Keeping track of the frequencies of graphlets during an iterative process proved difficult at first. 
\texttt{incgraph} provides a solution to this problem (Chapter~\ref{chap:incgraph}) by making assuming that the incremental changes between two different networks are relatively small (less than 100 edge additions/removals)). Under this assumption, it is possible to calculate the differences in topological patterns significantly faster in comparison to recomputing the graphlet frequencies from scratch at every iteration.

>>>>

We developed \texttt{dyngen}, a simulator of single cells to benchmark and stress-test computational tools for single-cell omics (Chapter~\ref{chap:dyngen}).
The \texttt{dyngen} software supports a selection of existing experimental protocols, and can also simulate experiments that are not possible with the current technologies, such as generating omics profiles from the same cell at multiple timepoints.
The generated data can help kick-start emerging domains with low data availability more safely by allowing software developers to test their method before numerous datasets become publicly available.

@@

The topological 'profile' of a network was defined as the frequencies of particular topological patterns called 'graphlets'. Keeping track of the frequencies of graphlets during an iterative process proved difficult at first. 
\texttt{incgraph} provides a solution to this problem (Chapter~\ref{chap:incgraph}) by  assuming that the incremental changes between two different networks are relatively small (less than 100 edge additions/removals)). Under this assumption, it is possible to calculate the differences in topological patterns significantly faster in comparison to recomputing the graphlet frequencies from scratch at every iteration.

### 2019-11-20 ch1_introduction/content.tex
update

Specific examples of omics technologies are whole genome sequencing to determine the DNA sequence of an organism, and RNA sequencing to profiles the sequence of RNA transcripts, both using next-generation sequencing technologies. A gene expression profile can be obtained by mapping the sequences of RNA transcripts to the genome.

@@

In this work, unless noted otherwise, we will be working with transcriptomics data resulting from a single-cell RNA-sequencing experiment (scRNA-seq). The workflow of generating scRNA-seq profiles is as follows. Same as other single-cell (non-omics) profiling methods, cells first need to be isolated (Figure~\ref{fig:profiling_convergence}B). Different sampling techniques yield different levels of information about cellular state. By now, many protocols for extracting and tagging RNA from single cells have been developed \cite{svensson_exponentialscalingsinglecell_2018}, the most popular of which are based on microfluidics or droplets (Figure~\ref{fig:profiling_convergence}C). By sequencing the transcripts and the attached unique cell identifier tags, each read can be mapped and tallied up. scRNA-seq data can thus be summarised in a matrix, where each column represent a single cell, each row a gene, and each value represents the number of transcripts that were sequenced for that gene and cell.

@@

Dimensionality reduction (DR) methods transform high-dimensional data into a meaningful low-dimensional representation. DR methods have two main target audiences; computers -- to construct a K-dimensional (with K $\ll$ M) representation of the data such that pairwise distances between different samples are retained as well as possible (Figure~\ref{fig:comp_tools_dimred}B); and humans -- to obtain a visual overview of the data (Figure~\ref{fig:comp_tools_dimred}C).

@@

\subsection{Differential expression}
Given that cells are split up two or more groups, differential expression (DE) methods ranks genes based on whether their expression is significantly higher or lower in one group in comparison to the others. This grouping can be based on prior information or an upstream clustering method. DE methods are useful for summarising the main differences between different groups of cells more compactly (Figure~\ref{fig:comp_tools_de}A) in comparison to when groups are compared without gene prioritisation (Figure~\ref{fig:comp_tools_cluster}A).

>>>>

Specific examples of omics technologies are whole genome sequencing to determine the DNA sequence of an organism, and RNA sequencing to profile the sequence of RNA transcripts, both using next-generation sequencing technologies. A gene expression profile can be obtained by mapping the sequences of RNA transcripts to the genome.

@@

In this work, unless noted otherwise, we will be working with transcriptomics data resulting from a single-cell RNA-sequencing experiment (scRNA-seq). The workflow of generating scRNA-seq profiles is as follows. Same as other single-cell (non-omics) profiling methods, cells first need to be isolated (Figure~\ref{fig:profiling_convergence}B). Different sampling techniques yield different levels of information about cellular state. By now, many protocols for extracting and tagging RNA from single cells have been developed \cite{svensson_exponentialscalingsinglecell_2018}, the most popular of which are based on microfluidics or droplets (Figure~\ref{fig:profiling_convergence}C). By sequencing the transcripts and the attached unique cell identifier tags, each read can be mapped and tallied up. scRNA-seq data can thus be summarised in a matrix, where each column represents a single cell, each row a gene, and each value represents the number of transcripts that were sequenced for that gene and cell.

@@

Dimensionality reduction (DR) methods transform high-dimensional data into a meaningful low-dimensional representation. DR methods have two main target audiences; computers -- to construct a K-dimensional (with K $\ll$ N) representation of the data such that pairwise distances between different samples are retained as well as possible (Figure~\ref{fig:comp_tools_dimred}B); and humans -- to obtain a visual overview of the data (Figure~\ref{fig:comp_tools_dimred}C).

@@

\subsection{Differential expression}
Given that cells are split up two or more groups, differential expression (DE) methods rank genes based on whether their expression is significantly higher or lower in one group in comparison to the others. This grouping can be based on prior information or an upstream clustering method. DE methods are useful for summarising the main differences between different groups of cells more compactly (Figure~\ref{fig:comp_tools_de}A) in comparison to when groups are compared without gene prioritisation (Figure~\ref{fig:comp_tools_cluster}A).
